Distributional Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain

### 2018-10-03 pkg/disttree/inst/draft/draft.Rnw
revisions based on reviews
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2306 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: d9572df8a9de22eab2f427c279fa463032ea6162

lines(x = x, y = y2, type = "l", col = pal[3], lwd = 1.3)
lines(x = x, y = y3, type = "l", col = pal[2], lwd = 1.3)
lines(x = x, y = y4, type = "l", col = pal[4], lwd = 1.3)
legend("topright", c("Predicted distribution", "Point mass at censoring point", "Observation"),
       bty = "n", col = "black", lty = c(1, NA, NA), pch = c(NA, 19, 4), cex = 0.8, lwd = 1.3)

@@

# add labels
text(x = -1.7, y = lh1, labels = "2009", col = pal[1], cex = 0.8)     # -0.8
text(x = -1.7, y = lh2, labels = "2010", col = pal[3], cex = 0.8)     # -0.8
text(x = -1.7, y = lh3, labels = "2011", col = pal[2], cex = 0.8)     # -0.8
text(x = -1.7, y = lh4, labels = "2012", col = pal[4], cex = 0.8)     # -0.8
@
\caption{\label{fig:axams-dist}\hl{Total precipitation predictions by a distributional forest
at station Axams for July 24 in 2009, 2010, 2011 and 2012 learned on data from 1985--2008.
Observations are modelled by a left-censored at 0 distribution. The corresponding predicted point mass is shown at the censoring point (0).}}
\end{figure}

@@

To incorporate the covariates into the distributional model, 
they are considered as regressors in additive predictors
$g_j(\theta_j) = f_{j,1}(\bold{Z}) + f_{j,2}(\bold{Z}) + \dots$
in GAMLSS.
Link functions $g_j(\cdot)$ are used for every parameter $\theta_j$
($j = 1, \dots, k$) based on smooth terms $f_{j,k}$ such as
nonlinear effects, spatial effects, random coefficients, or interaction surfaces
\citep{Klein+Kneib+Lang:2015}. However, this requires specifying the
additive terms and their functional forms in advance which can be challenging in
practice and potentially require domain knowledge, especially if the number of covariates $m$ is large.

@@

While Breiman and Cutler's random forests \citep{Breiman:2001} grow ensembles of trees
that pick up changes in the location of the response across the covariates,
\emph{distributional forests} employ an ensemble of $T$~\emph{distributional trees}. These
pick up changes in the ``direction'' of any distribution parameter by
considering the full score vector for choosing splitting variables and split
points. Each of the distributional trees is grown on a different data set
obtained through resampling (bootstrap or subsampling) and in each node only a
random subset of the covariates $\bold{Z}$ is considered. As usual in random
forests, this reduces the correlation among the trees and stabilizes the
variance of the model.

@@

\begin{figure}[t!]
\centering
\setkeys{Gin}{width=0.99\textwidth}
<<rain_axams_qqr_oosample, fig=TRUE, echo=FALSE, width = 6.5, pdf = FALSE, png = TRUE, resolution = 150>>=
# out of sample
set.seed(7)
par(mfrow = c(2, 2))
qqrplot(pit_df, nsim = 100, main = "Distributional forest", ylim = c(-5, 5), col = gray(0.04, alpha = 0.01), pch = 19)
qqrplot(pit_ml, nsim = 100, main = "EMOS",                  ylim = c(-5, 5), col = gray(0.04, alpha = 0.01), pch = 19)
qqrplot(pit_g,  nsim = 100, main = "Prespecified GAMLSS",   ylim = c(-5, 5), col = gray(0.04, alpha = 0.01), pch = 19)
qqrplot(pit_gb, nsim = 100, main = "Boosted GAMLSS",        ylim = c(-5, 5), col = gray(0.04, alpha = 0.01), pch = 19)
@
\caption{\label{fig:axams-qq} Out-of-sample residual QQ~plots (2009--2012) 
for station Axams based on models learned on data from 1985--2008. 
(Residuals of precipitation are on the powertransformed scale.)}
\end{figure}

@@

To assess calibration Figure~\ref{fig:axams-qq} shows residual QQ~plots for
out-of-sample predictions (2009--2012) from the different models trained on
1985--2008. Due to the point masses at zero 100 draws from the randomized
quantile residuals \citep{Dunn+Smyth:1996} are plotted in semi-transparent
gray. Overall, the randomized quantile residuals conform quite well with the
theoretical standard normal quantile (i.e., form a straight line close to the 
diagonal), indicating that all four models are sufficiently well calibrated. 
This is also supported by the corresponding probability integral transform
(PIT, \citealt{Gneiting+Balabdaoui+Raftery:2007}) histograms in
Appendix~\ref{app:pit}.

@@

\begin{figure}[t!]
\centering
\setkeys{Gin}{width=0.7\textwidth}
<<rain_axams_varim, fig=TRUE, echo=FALSE, width=6.5, height=4.2>>=
par(mar = c(2.5,10,1,2))
barplot(sort(vimp_crps, decreasing = FALSE)[(length(vimp_crps)-9):length(vimp_crps)], 
        horiz = TRUE, las = 1, axes = FALSE,
        xlab = "Variable importance: mean decrease in CRPS",
        font.axis = 3, #list(family="HersheySerif", face=3),
        names.arg = gsub("pow", "", names(sort(vimp_crps, decreasing = FALSE)[(length(vimp_crps)-9):length(vimp_crps)])))
axis(1, at = seq(0,1.6,0.2), las = 1, mgp=c(0,1,0))
@
\caption{\label{fig:axams-varimp} CRPS-based variable importance for the top 10 covariates 
in the distributional forest. Based on data for station Axams, learning period 
1985--2008 and assessed in 2009--2012.}
\end{figure}

>>>>

lines(x = x, y = y2, type = "l", col = pal[3], lwd = 1.3)
lines(x = x, y = y3, type = "l", col = pal[2], lwd = 1.3)
lines(x = x, y = y4, type = "l", col = pal[4], lwd = 1.3)
legend("topright", c("Predicted distribution", "Predicted point mass at 0", "Observation"),
       bty = "n", col = "black", lty = c(1, NA, NA), pch = c(NA, 19, 4), cex = 0.8, lwd = 1.3)

@@

# add labels
text(x = -1.7, y = lh1, labels = "2009", col = pal[1], cex = 0.8)     # -0.8
text(x = -1.7, y = lh2, labels = "2010", col = pal[3], cex = 0.8)     # -0.8
text(x = -1.7, y = lh3, labels = "2011", col = pal[2], cex = 0.8)     # -0.8
text(x = -1.7, y = lh4, labels = "2012", col = pal[4], cex = 0.8)     # -0.8
@
\caption{\label{fig:axams-dist}\hl{Total precipitation predictions by a 
distributional forest at station Axams for July 24 in 2009, 2010, 2011 and 
2012 learned on data from 1985--2008. Observations are non-negative and 
modeled by a Gaussian distribution, left-censored at 0.
The observations are depicted by crosses and the predicted point mass
from the model by filled circles.}}
\end{figure}

@@

To incorporate the covariates into the distributional model, 
they are considered as regressors in additive predictors
$g_j(\theta_j) = f_{j,1}(\bold{Z}) + f_{j,2}(\bold{Z}) + \dots$
in GAMLSS.
Link functions $g_j(\cdot)$ are used for every parameter $\theta_j$
($j = 1, \dots, k$) based on smooth terms $f_{j,k}$ such as
nonlinear effects, spatial effects, random coefficients, or interaction surfaces
\citep{Klein+Kneib+Lang:2015}. However, this requires specifying the
additive terms and their functional forms in advance which can be challenging in
practice and potentially require expert knowledge in the application domain, 
especially if the number of covariates $m$ is large.

@@

While Breiman and Cutler's random forests \citep{Breiman:2001} grow ensembles of trees
that pick up changes in the location of the response across the covariates,
\emph{distributional forests} employ an ensemble of $T$~\emph{distributional trees}. 
These pick up changes in the ``direction'' of any distribution parameter by
considering the full score vector for choosing splitting variables and split
points. \hl{For each distributional tree a different learning data set is obtained} 
through resampling (bootstrap or subsampling) and in each node only a
random subset of the covariates $\bold{Z}$ is considered. As usual in random
forests, this reduces the correlation among the trees and stabilizes the
variance of the model.

@@

\begin{figure}[t!]
\centering
\setkeys{Gin}{width=0.99\textwidth}
<<rain_axams_qqr_oosample, fig=TRUE, echo=FALSE, width = 6.5, pdf = FALSE, png = TRUE, resolution = 150>>=
# out of sample
set.seed(7)
par(mfrow = c(2, 2))
qqrplot(pit_df, nsim = 100, main = "Distributional forest", ylim = c(-5, 5), col = gray(0.04, alpha = 0.01), pch = 19)
qqrplot(pit_ml, nsim = 100, main = "EMOS",                  ylim = c(-5, 5), col = gray(0.04, alpha = 0.01), pch = 19)
qqrplot(pit_g,  nsim = 100, main = "Prespecified GAMLSS",   ylim = c(-5, 5), col = gray(0.04, alpha = 0.01), pch = 19)
qqrplot(pit_gb, nsim = 100, main = "Boosted GAMLSS",        ylim = c(-5, 5), col = gray(0.04, alpha = 0.01), pch = 19)
@
\caption{\label{fig:axams-qq} Out-of-sample residual QQ~plots (2009--2012) 
for station Axams based on models learned on data from 1985--2008.}
\end{figure}

@@

To assess calibration Figure~\ref{fig:axams-qq} shows residual QQ~plots for
out-of-sample predictions (2009--2012) from the different models trained on
1985--2008. Due to the point masses at zero 100 draws from the randomized
quantile residuals \citep{Dunn+Smyth:1996} are plotted in semi-transparent
gray. Overall, the randomized quantile residuals conform quite well with the
theoretical standard normal quantile (i.e., form a straight line close to the 
diagonal), indicating that all four models are sufficiently well calibrated. 
\hl{This is also supported by the corresponding probability integral transform}
(PIT, \citealt{Gneiting+Balabdaoui+Raftery:2007}) \hl{histograms for station
Axams in Supplement 2 (Stationwise Evaluation) which contains a more 
detailed explanation of residual QQ plots and PIT histograms.}

@@

\begin{figure}[t!]
\centering
\setkeys{Gin}{width=0.7\textwidth}
<<rain_axams_varim, fig=TRUE, echo=FALSE, width=6.5, height=4.2>>=
par(mar = c(2.5,10,1,2))
barplot(sort(vimp_crps, decreasing = FALSE)[(length(vimp_crps)-9):length(vimp_crps)], 
        horiz = TRUE, las = 1, axes = FALSE,
        xlab = "Variable importance: mean decrease in CRPS",
        font.axis = 3, #list(family="HersheySerif", face=3),
        xlim = c(0,1.1),
        names.arg = gsub("pow", "", names(sort(vimp_crps, decreasing = FALSE)[(length(vimp_crps)-9):length(vimp_crps)])))
axis(1, at = seq(0,1.6,0.2), las = 1, mgp=c(0,1,0))
@
\caption{\label{fig:axams-varimp} CRPS-based variable importance for the top 10 covariates 
in the distributional forest. Based on data for station Axams, learning period 
1985--2008 and assessed in 2009--2012.}
\end{figure}
